---
knitr:
  opts_chunk: 
    collapse: true
    cache: true
    message: false
editor: visual
---

```{r}
#| label: setup
#| echo: false
#| output: false

library(tidycensus)
library(tidyverse)
library(reactable)
library(sf)
library(arrow)
library(here)
library(htmlwidgets)
library(htmltools)
library(knitr)


options(tigris_use_cache = TRUE)
```

# Census Bureau data

:::::: panel-tabset
## American Community Survey 5-Year Data (2020-2024)

The ACS 5-year data for 2020-2024 was released on 01-29-2026^[1](https://www.census.gov/programs-surveys/acs/news/data-releases/2024/release-schedule.html)^. The interactive table below allows you to easily search through available datasets. Values from the **'Dataset ID'** column will be used to query the U.S. Census Bureau API\* in subsequent code chunks.

::: {style="font-size: 0.85em;"}
\**Note*: Census Bureau API keys can be requested [here](https://api.census.gov/data/key_signup.html). If you are trying to replicate this code and are running into issues, double check that you have an API key from the Census Bureau and that it has been added to your .Renviron by running the following:

|           \> tidycensus::census_api_key(key = \[**YOUR_API_KEY**\]))

R should implicitly reference this when making tidycensus api calls, but for added assurance, you can add your api key to an \_environment.local file.
:::

::: {.content-visible when-format="html"}
```{r}
#| label: acs_data
#| output: true

## Get 2024 ACS 5-year estimates
acs24 <- load_variables(2024, "acs5") |> 
  relocate(c("name", "concept"), .before = "label")


## create function for dropdown filter to use in reactable
filter_fct <- function(values, name) {
  tags$select(
    tags$option(value = "", "All"),
    purrr::map(unique(values), tags$option),
    onchange = glue::glue(
      "Reactable.setFilter(
        'my-tbl', // Modify id again
        '{name}', 
        event.target.value  
      )"
    ),
    style = "width: 100%; height: 28px;" # Add this line
  )
}

## Generate interactive table
reactable(acs24, 
          defaultColDef = colDef(
            header = function(value) gsub(".", " ", value, fixed = TRUE),
            cell = function(value) format(value, nsmall = 1),
            align = "left",
            headerStyle = list(background = "#F1F3F4")),
          theme = reactableTheme(
            highlightColor = "#D7E5F0",
            style = list(fontFamily = "-apple-system, BlinkMacSystemFont, Segoe UI, Helvetica, Arial, sans-serif"),
            searchInputStyle = list(width = "100%")),
          groupBy = "concept", 
          columns = list(
            name = colDef(name = "Dataset ID",
                          sticky = "left",
                          filterInput = filter_fct),
            concept = colDef(name = "Dataset name",
                             sticky = "left",
                             filterInput = filter_fct,
                             minWidth = 200),
            label = colDef(name = "Description",
                           minWidth = 100),
            geography = colDef(name = "Spatial scale",
                               aggregate = "unique", 
                               filterInput = filter_fct)),
          filterable = TRUE,
          highlight = TRUE,
          elementId = "my-tbl",
          searchable = TRUE,
          compact = TRUE,
          width = "100%",
          paginationType = "jump",
          minRows = 10,
          showPageSizeOptions = TRUE
          )

```
:::

## Decennial Census (2020)

Demographic and Housing Characteristics file. Explored these data to see if maybe there was more specific housing data available.

::: {.content-visible when-format="html"}
```{r}
#| label: dhc_data
#| output: true

dhc20 <- load_variables(2020, "dhc") |> 
  relocate(c("name", "concept"), .before = "label")


## create function for dropdown filter to use in reactable
filter_fct <- function(values, name) {
  tags$select(
    tags$option(value = "", "All"),
    purrr::map(unique(values), tags$option),
    onchange = glue::glue(
      "Reactable.setFilter(
        'my-tbl2', // Modify id again
        '{name}', 
        event.target.value  
      )"
    ),
    style = "width: 100%; height: 28px;" # Add this line
  )
}

## Generate interactive table
reactable(dhc20, 
          defaultColDef = colDef(
            header = function(value) gsub(".", " ", value, fixed = TRUE),
            cell = function(value) format(value, nsmall = 1),
            align = "left",
            headerStyle = list(background = "#F1F3F4")),
          theme = reactableTheme(
            highlightColor = "#D7E5F0",
            style = list(fontFamily = "-apple-system, BlinkMacSystemFont, Segoe UI, Helvetica, Arial, sans-serif"),
            searchInputStyle = list(width = "100%")),
          groupBy = "concept", 
          columns = list(
            name = colDef(name = "Dataset ID",
                          sticky = "left",
                          filterInput = filter_fct),
            concept = colDef(name = "Dataset name",
                             sticky = "left",
                             filterInput = filter_fct,
                             minWidth = 200),
            label = colDef(name = "Description",
                           minWidth = 200)),
          filterable = TRUE,
          highlight = TRUE,
          elementId = "my-tbl2",
          searchable = TRUE,
          compact = TRUE,
          width = "100%",
          paginationType = "jump",
          minRows = 10,
          showPageSizeOptions = TRUE
          )
```
:::
::::::

# Socioeconomic variables of interest

From searching the tables above, the following datasets were chosen:

## Race/ethnicity (B02001; Race \| census tract)

```{r}
#| label: race_and_ethnicity
#| output: true

## List of states to iterate through
states <- state.name

## Create labels for dataset variables
col_names <- acs24 |> 
  filter(concept == "Race") |> 
  separate_wider_delim(cols = label, 
                       delim = "!!", 
                       names = c("estimate", "var", "var_label"),
                       too_few = "align_end", too_many = "merge") |> 
  mutate(var_label = case_when(var_label == "Total:" ~ "Total",
                           var_label == "Two or More Races:" ~ "Two or More Races",
                           var_label == "Two or More Races:!!Two races including Some Other Race" ~ "Two races including Some Other Race",
                           var_label == "Two or More Races:!!Two races excluding Some Other Race, and three or more races" ~ "Two races excluding Some Other Race, and three or more races",
                           TRUE ~ var_label)) |> 
  subset(select = -c(concept, estimate, var, geography)) |> 
  rename(variable = name)

dict <- setNames(col_names$var_label, col_names$variable)


## Loop through each state and merge the tract data
race <- map_dfr(states, function(st) {
  get_acs(geography = "tract",
          table = "B02001", # Race
          state = st,
          year = 2024,
          summary_var = "B02001_001",
          geometry = T, 
          keep_geo_vars = T
          )
  },
  .progress = T) |> 
  # remove redundant vars
  dplyr::select(-c(NAME.x, NAME.y, GEOIDFQ, LSAD, STUSPS)) |> 
  # label intuitively
  rename(state_fips = STATEFP,
         county_fips = COUNTYFP,
         census_tract_fips = TRACTCE,
         census_tract = NAMELSAD,
         county = NAMELSADCO,
         state = STATE_NAME,
         total_est = summary_est,
         total_moe = summary_moe) |>
  relocate(GEOID, .before = state_fips) |> 
  relocate(c(state, county), .before = census_tract) |> 
  relocate(c(ALAND, AWATER), .before = geometry) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable) |> 
  filter(var_label != "Total")


## Save data in .gpkg and as parquet file -- commented out here to run code more easily
# st_write(race, here("02_outputs", "socioeconomic_data.gpkg"), layer = "race_B02001")

# st_drop_geometry(race) |> 
#   write_parquet(here("02_outputs", "race_all.parquet"))

## Save CA and IN separately
# st_drop_geometry(race) |>
#   filter(state == "Indiana") |>
#   write_parquet(here("02_outputs", "race_IN.parquet"))
# 
# st_drop_geometry(race) |>
#   filter(state == "California") |>
#   write_parquet(here("02_outputs", "race_CA.parquet"))

```

- Data for the whole U.S. at the tract level is available as a .gpkg and a parquet file.

#### \> Consolidate groups

```{r}
#| label: race_simplified

race_simplified <- race |> 
  mutate(var_label = case_when(var_label == "White alone" ~ "White",
                               var_label == "Black or African American alone" ~ "Black",
                               var_label == "American Indian and Alaska Native alone" ~ "Indigenous",
                               var_label == "Asian alone" ~ "Asian",
                               TRUE ~ "Other")) |>
  dplyr::select(-(variable)) |> 
  group_by(census_tract, var_label) |> 
  
  # Pretty sure this isn't a great way of doing this but let me know. Need to figure something else out.
  mutate(estimate = mean(estimate),
         moe = mean(moe),
         total_est = mean(total_est),
         total_moe =  mean(total_moe),
         ALAND = mean(ALAND),
         AWATER = mean(AWATER)) 


# |> 
#   distinct()

rm(race) # added this to free up memory


## In progress -- ignore
# indiana_cali <- race_simplified |> 
#   filter(state == "Indiana" | state == "California") |> 
#   group_by(census_tract, var_label) |> 
#   mutate(estimate = mean(estimate),
#          moe = mean(moe),
#          total_est = mean(total_est),
#          total_moe =  mean(total_moe),
#          ALAND = mean(ALAND),
#          AWATER = mean(AWATER)) |> 
#   distinct() |> 
#   mutate(proportion = round(estimate/total_est), 2)

rm(race, race_simplified, col_names, dict)
gc()

```

## Income (B19013_001; Income \| census tract)

```{r}
#| label: median_income
#| output: true

col_names <- acs24 |> 
  filter(name == "B19013_001") |> 
  rename(variable = name,
         var_label = concept)

dict <- setNames(col_names$var_label, col_names$variable)

## Loop through each state and merge the tract data
income <- map_dfr(states, function(st) {
  get_acs(geography = "tract",
          table = "B19013", # Race
          state = st,
          year = 2024,
          geometry = T, 
          keep_geo_vars = T
          )
  },
  .progress = T) |> 
  # remove redundant vars
  dplyr::select(-c(NAME.x, NAME.y, GEOIDFQ, LSAD, STUSPS)) |> 
  # label intuitively
  rename(state_fips = STATEFP,
         county_fips = COUNTYFP,
         census_tract_fips = TRACTCE,
         census_tract = NAMELSAD,
         county = NAMELSADCO,
         state = STATE_NAME) |>
  relocate(GEOID, .before = state_fips) |> 
  relocate(c(state, county), .before = census_tract) |> 
  relocate(c(ALAND, AWATER), .before = geometry) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable) 


# ## Save data in .gpkg and as parquet file -- commented out here to run code more easily
# st_write(income, here("02_outputs", "socioeconomic_data.gpkg"), layer = "income_B19013", append = T)
# 
# st_drop_geometry(income) |>
# write_parquet(here("02_outputs", "income_all.parquet"))
# 
# ## Save CA and IN separately
# st_drop_geometry(income) |>
#   filter(state == "Indiana") |>
#   write_parquet(here("02_outputs", "income_IN.parquet"))
# 
# st_drop_geometry(income) |>
#   filter(state == "California") |>
#   write_parquet(here("02_outputs", "income_CA.parquet"))


rm(income, col_names, dict)
gc()
```

## Median age (B01002; Median age by sex \| block group)

\**Note to self*: add code generalization to other snippets starting here

```{r}
#| label: median_age_by_sex
#| output: true

## Median age by sex
col_names <- acs24 |> 
  filter(str_detect(name, "B01002_")) |> 
  separate_wider_delim(cols = label, 
                       delim = "!!", 
                       names = c("estimate", "var", "var_label"),
                       too_few = "align_end", too_many = "merge") |> 
  mutate(var_label = case_when(var_label == "Total:" ~ "Total",
                               TRUE ~ var_label)) |> 
  subset(select = -c(concept, estimate, var, geography)) |> 
  rename(variable = name)

  
dict <- setNames(col_names$var_label, col_names$variable)

## Get ACS dataset of interest
age_IN <- get_acs(geography = "block group",
                table = "B01002",
                year = 2024,
                state = "Indiana",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)

age_CA <- get_acs(geography = "block group",
                table = "B01002",
                year = 2024,
                state = "California",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)


# write_parquet(age_IN, here("02_outputs", "age_IN.parquet"))
# write_parquet(age_CA, here("02_outputs", "age_CA.parquet"))


rm(age_CA, age_IN, col_names, dict)
gc()
```

## Housing Units (B25001_001; Housing Units \| block group)

Another potential dataset is **P16_001N-P16_009N** (Household type) from the Demographic History Sample (DHS) portion of the 2020 Decennial Survey (i.e., variables are things like 'married couple family'; 'non-family households' etc.). If we are looking for data more similar to building type, that exists already in our NSI datasets.

***Please let me know if I need to get a different dataset for this.***

```{r}
#| label: housing_units
#| output: true

col_names <- acs24 |> 
  filter(str_detect(name, "B25001_")) |>  
  rename(variable = name,
         var_label = concept)

  
dict <- setNames(col_names$var_label, col_names$variable)

## Get ACS dataset of interest
housing_units_IN <- get_acs(geography = "block group",
                table = "B25001",
                year = 2024,
                state = "Indiana",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)

housing_units_CA <- get_acs(geography = "block group",
                table = "B25001",
                year = 2024,
                state = "California",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)


# write_parquet(housing_units_IN, here("02_outputs", "housing_units_IN.parquet"))
# write_parquet(housing_units_CA, here("02_outputs", "housing_units_CA.parquet"))


rm(housing_units_CA, housing_units_IN, col_names, dict)
gc()


```

## Population density (P1_001N; Total Pop \| census tract)

- I would like clarification on how we want pop. density to be presented

```{r}
#| label: population_density
#| output: true

## total population by census tract
col_names <- dhc20 |> 
  filter(str_detect(name, "P1_001N")) |> 
  mutate(var_label = str_to_title(concept)) |> 
  subset(select = -c(concept, label)) |> 
  rename(variable = name)

  
dict <- setNames(col_names$var_label, col_names$variable)


## Get total population by block for a state
total_pop <- get_decennial(
  geography = "tract",
  variables = "P1_001N", # Total population variable (2020 Census)
  state = "Indiana",      
  year = 2020,
  geometry = TRUE,
  keep_geo_vars = TRUE) 

glimpse(total_pop)

pop_density <-  total_pop |> 
##  Convert ALAND (Square Meters) to sq km or sq mi
  mutate(land_area_sq_km = ALAND / 1000000,
         land_area_sq_mile = ALAND / 2589988) |> 
## Calculate population density 
  mutate(pop_density_per_sq_km = value / land_area_sq_km,
         pop_density_per_sq_mile = value / land_area_sq_mile) |> 
  relocate(c(land_area_sq_km, pop_density_per_sq_km), .after = ALAND) |> 
  relocate(c(land_area_sq_km, pop_density_per_sq_km), .after = ALAND) |>
  # remove redundant vars
  dplyr::select(-c(NAME.x, NAME.y, AFFGEOID, LSAD, STUSPS)) |> 
  # label intuitively
  rename(state_fips = STATEFP,
         county_fips = COUNTYFP,
         census_tract_fips = TRACTCE,
         census_tract = NAMELSAD,
         county = NAMELSADCO,
         state = STATE_NAME) |>
  relocate(GEOID, .before = state_fips) |> 
  relocate(c(state, county), .before = census_tract) |> 
  relocate(c(ALAND, AWATER), .before = geometry) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable) 
  
  
head(pop_density)


## Save data in .gpkg and as parquet file -- commented out here to run code more easily
st_write(pop_density, here("02_outputs", "socioeconomic_data.gpkg"), layer = "pop_density_P1_001N")

# st_drop_geometry(pop_density) |> 
#   write_parquet(here("02_outputs", "pop_density_all.parquet"))

# Save CA and IN separately
# st_drop_geometry(pop_density) |>
#   filter(state == "Indiana") |>
#   write_parquet(here("02_outputs", "pop_density_IN.parquet"))
# 
# st_drop_geometry(pop_density) |>
#   filter(state == "California") |>
#   write_parquet(here("02_outputs", "pop_density_CA.parquet"))



```

## Food deserts

Data related to food deserts were downloaded directly from the USDA's website^[2](https://www.ers.usda.gov/data-products/food-access-research-atlas/download-the-data)^. This dataset - 2019 Large Retailer Access Map (LRAM) (formerly known as the Food Access Research Atlas (FARA)) - was last updated 04-27-2021.

```{r}
#| label: food_deserts
#| output: true
#| warning: false

FARA_all <- read.csv(here("01_data/2019_Food_Access_Research_Atlas_Data", "FARA.csv"), header = T, encoding = "UTF-8")
var_lookup <- read.csv(here("01_data/2019_Food_Access_Research_Atlas_Data", "VariableLookup.csv")) 

FARA_subset <- FARA_all |> 
  subset(select = c("CensusTract", "County", "State", "Urban", "LowIncomeTracts", "PovertyRate", "LAPOP1_10",
                    "LAPOP05_10", "LAPOP1_20", "lapop1", "lapophalf", "lapop10", "TractHUNV", "lahunv1", 
                    "lahunvhalf", "lahunv10", "lahunv20", "LATractsVehicle_20", "LALOWI1_10", "LALOWI05_10", 
                    "LALOWI1_20", "lalowi1", "lalowihalf", "lalowi10", "lalowi20")) |> 
  rename(census_tract = CensusTract,
         county = County, 
         state = State) |> 
  mutate(across(c(4:5, 18), as.factor)) |>
  mutate(across(c(6:17, 19:25), as.integer))
              
FARA_IN <- FARA_subset |> 
  filter(state == "Indiana")

FARA_CA <- FARA_subset |> 
  filter(state == "California")

write_parquet(FARA_IN, here("02_outputs", "FARA_IN.parquet"))
write_parquet(FARA_CA, here("02_outputs", "FARA_CA.parquet"))


```

|  A lot of this data overlaps with ACS data (e.g. race, median income, age, etc.), although I think it may be slightly more outdated. For example, they have census tract population numbers from 2010. While I think it's best for us to use data from the same dataset when we can, I'm wondering if there is a way to balance this food access data with more up-to-date income, age, and population data from ACS without totally biasing analyses.
| For now, I am just subsetting the portions of the dataset that have to do with access.

## Green space

I found a few data sources for this, but would like opinions on what I should use:

1.  [UN Urban Indicators Database](https://data.unhabitat.org/pages/open-spaces-and-green-areas)

2.  [This table of datasets from Georgetown University](https://mdi.georgetown.edu/pbi/greenspace/)

3.  [Green space per capita (m^2^/person) from the EPA](https://www.epa.gov/enviroatlas/enviroatlas-dynamic-data-matrix)



## Other thoughts
- Please let me know what you think of my choice in dataset from the census bureau.
- We want to make sure we have data at the *census tract* spatial scale right?

