---
knitr:
  opts_chunk: 
    collapse: true
    cache: true
editor: visual
---

```{r}
#| label: setup
#| echo: false
#| output: false

library(tidycensus)
library(tidyverse)
library(reactable)
library(sf)
library(here)
library(htmlwidgets)
library(htmltools)
library(knitr)


options(tigris_use_cache = TRUE)
```

## American Community Survey 5-Year Data (2020-2024)

The ACS 5-year data for 2020-2024 was released on 01-29-2026^[1](https://www.census.gov/programs-surveys/acs/news/data-releases/2024/release-schedule.html)^. The interactive table below allows you to easily search through available datasets. Values from the **'Dataset ID'** column will be used to query the U.S. Census Bureau API\* in subsequent code chunks.

::: {style="font-size: 0.85em;"}
\**Note*: Census Bureau API keys can be requested [here](https://api.census.gov/data/key_signup.html). If you are trying to replicate this code and are running into issues, double check that you have an API key from the Census Bureau and that it has been added to your .Renviron by running the following:

|           \> tidycensus::census_api_key(key = \[**YOUR_API_KEY**\]))

R should implicitly reference this when making tidycensus api calls, but for added assurance, you can add your api key to an \_environment.local file.
:::

::: {.content-visible when-format="html"}
```{r}
#| label: acs_data
#| output: true

## Get 2024 ACS 5-year estimates
acs24 <- load_variables(2024, "acs5") |> 
  relocate(c("name", "concept"), .before = "label")


## create function for dropdown filter to use in reactable
filter_fct <- function(values, name) {
  tags$select(
    tags$option(value = "", "All"),
    purrr::map(unique(values), tags$option),
    onchange = glue::glue(
      "Reactable.setFilter(
        'my-tbl', // Modify id again
        '{name}', 
        event.target.value  
      )"
    ),
    style = "width: 100%; height: 28px;" # Add this line
  )
}

## Generate interactive table
reactable(acs24, 
          defaultColDef = colDef(
            header = function(value) gsub(".", " ", value, fixed = TRUE),
            cell = function(value) format(value, nsmall = 1),
            align = "left",
            headerStyle = list(background = "#F1F3F4")),
          theme = reactableTheme(
            highlightColor = "#D7E5F0",
            style = list(fontFamily = "-apple-system, BlinkMacSystemFont, Segoe UI, Helvetica, Arial, sans-serif"),
            searchInputStyle = list(width = "100%")),
          groupBy = "concept", 
          columns = list(
            name = colDef(name = "Dataset ID",
                          sticky = "left",
                          filterInput = filter_fct),
            concept = colDef(name = "Dataset name",
                             sticky = "left",
                             filterInput = filter_fct,
                             minWidth = 200),
            label = colDef(name = "Description",
                           minWidth = 100),
            geography = colDef(name = "Spatial scale",
                               aggregate = "unique", 
                               filterInput = filter_fct)),
          filterable = TRUE,
          highlight = TRUE,
          elementId = "my-tbl",
          searchable = TRUE,
          compact = TRUE,
          width = "100%",
          paginationType = "jump",
          minRows = 10,
          showPageSizeOptions = TRUE
          )

```
:::

## Decennial Census (2020)

Demographic and Housing Characteristics file. Explored these data to see if maybe there was more specific housing data available.

::: {.content-visible when-format="html"}
```{r}
#| label: dhc_data
#| output: true

dhc20 <- load_variables(2020, "dhc") |> 
  relocate(c("name", "concept"), .before = "label")


## create function for dropdown filter to use in reactable
filter_fct <- function(values, name) {
  tags$select(
    tags$option(value = "", "All"),
    purrr::map(unique(values), tags$option),
    onchange = glue::glue(
      "Reactable.setFilter(
        'my-tbl2', // Modify id again
        '{name}', 
        event.target.value  
      )"
    ),
    style = "width: 100%; height: 28px;" # Add this line
  )
}

## Generate interactive table
reactable(dhc20, 
          defaultColDef = colDef(
            header = function(value) gsub(".", " ", value, fixed = TRUE),
            cell = function(value) format(value, nsmall = 1),
            align = "left",
            headerStyle = list(background = "#F1F3F4")),
          theme = reactableTheme(
            highlightColor = "#D7E5F0",
            style = list(fontFamily = "-apple-system, BlinkMacSystemFont, Segoe UI, Helvetica, Arial, sans-serif"),
            searchInputStyle = list(width = "100%")),
          groupBy = "concept", 
          columns = list(
            name = colDef(name = "Dataset ID",
                          sticky = "left",
                          filterInput = filter_fct),
            concept = colDef(name = "Dataset name",
                             sticky = "left",
                             filterInput = filter_fct,
                             minWidth = 200),
            label = colDef(name = "Description",
                           minWidth = 200)),
          filterable = TRUE,
          highlight = TRUE,
          elementId = "my-tbl2",
          searchable = TRUE,
          compact = TRUE,
          width = "100%",
          paginationType = "jump",
          minRows = 10,
          showPageSizeOptions = TRUE
          )
```
:::

## Socioeconomic variables of interest

From searching the tables above, the following datasets were chosen:

#::: panel-tabset
### Race/ethnicity

- B02001 (Race \| block group)


```{r}
#| label: race_and_ethnicity_TEST
#| output: false

## List of states to iterate through
states <- state.name

## Loop through each state and merge the tract data
race_data <- map_dfr(states, function(st) {
  get_acs(geography = "block group",
          table = "B02001", # Race
          state = st,
          year = 2024,
          summary_var = "B02001_001",
          geometry = T, 
          keep_geo_vars = T
          )
  },
  .progress = T)

## Create labels for dataset variables
col_names <- acs24 |> 
  filter(concept == "Race") |> 
  separate_wider_delim(cols = label, 
                       delim = "!!", 
                       names = c("estimate", "var", "var_label"),
                       too_few = "align_end", too_many = "merge") |> 
  mutate(var_label = case_when(var_label == "Total:" ~ "Total",
                           var_label == "Two or More Races:" ~ "Two or More Races",
                           var_label == "Two or More Races:!!Two races including Some Other Race" ~ "Two races including Some Other Race",
                           var_label == "Two or More Races:!!Two races excluding Some Other Race, and three or more races" ~ "Two races excluding Some Other Race, and three or more races",
                           TRUE ~ var_label)) |> 
  subset(select = -c(concept, estimate, var, geography)) |> 
  rename(variable = name)

dict <- setNames(col_names$var_label, col_names$variable)

race_data <- race_data |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable) |> 
  mutate(across(everything(), \(x) trimws(x, which = "both"))) |> 
  filter(var_label != "Total")


```

#### Summary Table

#### Map Overview
```{r}
race_data |>
  mutate(percent = 100 * (estimate / summary_estimate)) |>
  ggplot(aes(fill = percent, colour = )) +
  facet_wrap(~variable) +
  geom_sf(color = NA) +
  theme_void() + 
  scale_fill_viridis_c() + 
  labs(fill = "% of population\n(2020 Census)")
```

```{r}
#| label: race_and_ethnicity
#| output: true


## Get ACS dataset of interest
race_IN <- get_acs(geography = "block group",
                table = "B02001",
                year = 2024,
                state = "Indiana",
                survey = "acs5") 

race_CA <- get_acs(geography = "block group",
                table = "B02001",
                year = 2024,
                state = "California",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)


#dir <- file.path(dirname(rstudioapi::getActiveDocumentContext()$path))
#out_path <- file.path(dir, path.expand("outputs"))
#dir.create(out_path)

rm(col_names, dict)
```

### Income

- B19013_001 (Income \| block group)

```{r}
#| label: median_income
#| output: true

col_names <- acs24 |> 
  filter(name == "B19013_001") |> 
  rename(variable = name,
         var_label = concept)

dict <- setNames(col_names$var_label, col_names$variable)

## Get ACS dataset of interest
income_IN <- get_acs(geography = "block group",
                table = "B19013",
                year = 2024,
                state = "Indiana",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)

income_CA <- get_acs(geography = "block group",
                table = "B19013",
                year = 2024,
                state = "California",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)

rm(col_names, dict)
gc()
```

### Median age

- B01002 (Median age by sex \| block group)

```{r}
#| label: median_age_by_sex
#| output: true

## Median age by sex
col_names <- acs24 |> 
  filter(str_detect(name, "B01002_")) |> 
  separate_wider_delim(cols = label, 
                       delim = "!!", 
                       names = c("estimate", "var", "var_label"),
                       too_few = "align_end", too_many = "merge") |> 
  mutate(var_label = case_when(var_label == "Total:" ~ "Total",
                               TRUE ~ var_label)) |> 
  subset(select = -c(concept, estimate, var, geography)) |> 
  rename(variable = name)

  
dict <- setNames(col_names$var_label, col_names$variable)

## Get ACS dataset of interest
age_IN <- get_acs(geography = "block group",
                table = "B01002",
                year = 2024,
                state = "Indiana",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)

age_CA <- get_acs(geography = "block group",
                table = "B01002",
                year = 2024,
                state = "California",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)

rm(col_names, dict)
gc()
```

### Single family households

- B25001_001 (Housing Units \| block group)

Another potential dataset is **P16_001N-P16_009N** (Household type) from the Demographic History Sample (DHS) portion of the 2020 Decennial Survey (i.e., variables are things like 'married couple family'; 'non-family households' etc.). If we are looking for data more similar to building type, that exists already in our NSI datasets.

***Please let me know if I need to get a different dataset for this.***

```{r}
#| label: housing_units
#| output: true

col_names <- acs24 |> 
  filter(str_detect(name, "B25001_")) |>  
  rename(variable = name,
         var_label = concept)

  
dict <- setNames(col_names$var_label, col_names$variable)

## Get ACS dataset of interest
housing_units_IN <- get_acs(geography = "block group",
                table = "B25001",
                year = 2024,
                state = "Indiana",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)

housing_units_CA <- get_acs(geography = "block group",
                table = "B25001",
                year = 2024,
                state = "California",
                survey = "acs5") |> 
  separate_wider_delim(cols = NAME, 
                       delim = ";", 
                       names = c("block_group", "census_tract", "county", "state"),
                       too_few = "error", too_many = "error",
                       cols_remove = TRUE) |> 
  mutate(var_label = recode(variable, !!!dict)) |> 
  relocate(var_label, .after = variable)

rm(col_names, dict)
gc()
```

### Population density

```{r}
#| label: population_density
#| output: true


## Get total population by block for a state
total_pop <- get_decennial(
  geography = "tract",
  variables = "P1_001N", # Total population variable (2020 Census)
  state = "Indiana",      
  year = 2020,
  geometry = TRUE,
  keep_geo_vars = TRUE) 

glimpse(total_pop)

pop_density <-  total_pop |> 
##  Convert ALAND (Square Meters) to sq km or sq mi
  mutate(land_area_sq_km = ALAND / 1000000,
         land_area_sq_mile = ALAND / 2589988) |> 
## Calculate population density 
  mutate(pop_density_per_sq_km = value / land_area_sq_km,
         pop_density_per_sq_mile = value / land_area_sq_mile)
  
  
test <- pop_density


## total population by census tract
col_names <- acs24 |> 
  filter(str_detect(name, "B01003_")) #|> 
  separate_wider_delim(cols = label, 
                       delim = "!!", 
                       names = c("estimate", "var", "var_label"),
                       too_few = "align_end", too_many = "merge") |> 
  mutate(var_label = case_when(var_label == "Total:" ~ "Total",
                               TRUE ~ var_label)) |> 
  subset(select = -c(concept, estimate, var, geography)) |> 
  rename(variable = name)

  
dict <- setNames(col_names$var_label, col_names$variable)

############# TRYING SOMETHING HERE #############
ddhca_pop_groups <- get_pop_groups(2020, 
                                   sumfile = "ddhca")


```

### Food deserts

Data related to food deserts were downloaded directly from the USDA's website^[2](https://www.ers.usda.gov/data-products/food-access-research-atlas/download-the-data)^. This dataset - 2019 Large Retailer Access Map (LRAM) (formerly known as the Food Access Research Atlas (FARA)) - was last updated 04-27-2021.

```{r}
#| label: food_deserts
#| output: true
#| warning: false

FARA_all <- read.csv(here("01_data/2019_Food_Access_Research_Atlas_Data", "FARA.csv"), header = T, encoding = "UTF-8")
var_lookup <- read.csv(here("01_data/2019_Food_Access_Research_Atlas_Data", "VariableLookup.csv")) 

FARA_subset <- FARA_all |> 
  subset(select = c("CensusTract", "County", "State", "Urban", "LowIncomeTracts", "PovertyRate", "LAPOP1_10",
                    "LAPOP05_10", "LAPOP1_20", "lapop1", "lapophalf", "lapop10", "TractHUNV", "lahunv1", 
                    "lahunvhalf", "lahunv10", "lahunv20", "LATractsVehicle_20", "LALOWI1_10", "LALOWI05_10", 
                    "LALOWI1_20", "lalowi1", "lalowihalf", "lalowi10", "lalowi20")) |> 
  rename(census_tract = CensusTract,
         county = County, 
         state = State) |> 
  mutate(across(c(4:5, 18), as.factor)) |>
  mutate(across(c(6:17, 19:25), as.integer))
              
FARA_IN <- FARA_subset |> 
  filter(state == "Indiana")

FARA_CA <- FARA_subset |> 
  filter(state == "California")

```
|  A lot of this data overlaps with ACS data (e.g. race, median income, age, etc.), although I think it may be slightly more outdated. For example, they have census tract population numbers from 2010. While I think it's best for us to use data from the same dataset when we can, I'm wondering if there is a way to balance this food access data with more up-to-date income, age, and population data from ACS without totally biasing analyses.
| For now, I am just subsetting the portions of the dataset that have to do with access.


### Green space

```{r}

```
#:::
